%-----------------------------------------------------------------------------%
%- Acknowledgments -----------------------------------------------------------%
%-----------------------------------------------------------------------------%
\chapter{Acknowledgements}
\label{chap:Acknowledgements}
I had a whole swag of people to help me along the way. Listed, in no particular
order (because there is no fair way to order you), they are:
\begin{description}
  \item[Carl Svensson] Macquarie University, for the \LaTeX help, the maths, and
  the many late night sessions over a family dinner box, and the many in-jokes
  and innuendoes\footnote{Giggity}.
  
  \item[Michael Griffin, Josh Larietti, Gareth Richardson, Rajika Kuruwita,
  Justin Byrne] Macquarie University, for proof reading, finding errors, and
  being honest\footnote{blunt} when something could be improved\footnote{ie,
  if it was ``shit'', you'd call it ``shit'' ;)}, the many in-jokes, robot
  building, over-eating of junk food, maths, hijinx, sounding out of ideas and
  whatever else goes into peer assisted learning into all hours of the night.
  
  \item[The Heimlich Family] Macquarie University, for giving me a fantastic
  opportunity to put things I've learned into practice, for the life lessons
  that resulted from it. To Mike, Luan, Sarah and Jaye for making FIRST happen
  in Australia, and for inviting me to join the FIRST family.
  
  \item[Dom Verity] Macquarie University, for being a source of inspiration and
  convincing me to try\{\} just one more time before I threw an exception.
  
  \item[MQ Deptartment of Electronics Engineering staff]\footnote{Ana, without
  you, we could never get anything done} David Wong, Yinan, Sam, Tony, Rein,
  Barry, Gengfa, Dush, Hoque, Karu - for making engineering awesome!
  
  \item[MQ Department of Mathematics staff] In particular Fran Grifin, Chris
  Gordon, Rod Yager, Geoff, Dilshara Hill, and Daniel Linssen. You make maths
  comprehensible\footnote{for the most part!}.
  
  \item[FIRST Team 3132] The Thunder Down Under, for always holding me to high
  standards of Gracious Professionalism\texttrademark\footnote{Gracious
  Professionalism is a common law trademark of the United States Foundation for
  Inspiration and Recognition of Science and Technology (US FIRST).}
  
  \item[Mark Leon] NASA, for the words of inspiration and wisdom when you
  spoke at the 2011 Honolulu FIRST FRC regionals:
  ``\ldots at the end of the day, it will be the engineers who save the world
  \ldots This is why we do the math\ldots''
  
  \item[Donald Knuth] for giving me \LaTeX to typeset this book, The Art of
  Computer Programming and Concrete Mathematics as reference materials, as well
  as the awesomeness you've contributed to computing and maths in general.
  
  \item[Celeste Cohen] For letting me show off stuff to you which I thought
  was pretty cool, whilst boring you to tears, as well as advice on the page
  layout and wording. For the Millie. But above all, for being there. /me hugs
  you
  
  \item[Millie] 4 teh nyanz, teh meowz, teh nomz, teh sitz, teh warmz, teh
  purrrrz, and for\footnote{alwayz!} waking me up 45 minutez before teh
  alarm wuz due 2 go off, bcz u wantz 2 play wif u and teh jingly ball.
  
  \item[Pit Crew] it would be remiss of me to not mention the pit crew who make
  sure that I keep going lap after lap\ldots Nathan, Nick, Diana, Heidi, Hugh,
  The Gordon Family\footnote{Put a donk on it!}, Nadia, Martin and Jessica,
  Frankie, the hams at VK2BV, Stephen VK2TQ, Will, Pippa, David, Emily, Andrew,
  John, Sue, Matthew, Richard, my brother Sean, and my mother and my father.
\end{description}